Gaussian Process regression for astronomical time-series

### 2022-08-08 src/tex/ms.tex
[showyourwork] overleaf sync

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}

% Begin!
\begin{document}

% Title
\title{An open source scientific article}

% Author list
\author{@dfm}

% Abstract with filler text
\begin{abstract}
    Lorem ipsum dolor sit amet, consectetuer adipiscing elit. 
    Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis. 
    Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna. 
    Donec vehicula augue eu neque, morbi tristique senectus et netus et. 
    Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum. 
    Phasellus eu tellus sit amet tortor gravida placerat. 
    Integer sapien est, iaculis in, pretium quis, viverra ac, nunc. 
    Praesent eget sem vel leo ultrices bibendum. 
    Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac. 
    Curabitur auctor semper nulla donec varius orci eget risus. 
    Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam. 
    Duis eget orci sit amet orci dignissim rutrum.
\end{abstract}

% Main body with filler text
\section{Introduction}
\label{sec:intro}

Lorem ipsum dolor sit amet, consectetuer adipiscing elit. 
Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis. 
Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna. 
Donec vehicula augue eu neque, morbi tristique senectus et netus et. 
Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum. 
Phasellus eu tellus sit amet tortor gravida placerat. 
Integer sapien est, iaculis in, pretium quis, viverra ac, nunc. 
Praesent eget sem vel leo ultrices bibendum. 
Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac. 
Curabitur auctor semper nulla donec varius orci eget risus. 
Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam. 
Duis eget orci sit amet orci dignissim rutrum.

Nam dui ligula, fringilla a, euismod sodales, sollici- tudin vel, wisi. 
Morbi auctor lorem non justo, nam lacus libero, pretium at, lobortis vitae. 
Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna. 
Morbi ac orci et nisl hendrerit mollis, suspendisse ut massa, cras nec ante. 
Pellentesque a nulla cum sociis natoque penatibus et magnis dis parturient. 
Aliquam tincidunt urna, nulla ullamcorper vestibulum turpis. 
Pellentesque cursus luctus mauris \citep{Luger2021}.

\bibliography{bib}

>>>>

% template.tex, dated April 5 2013
% This is a template file for Annual Reviews 1 column Journals
%
% Compilation using ar-1col.cls' - version 1.0, Aptara Inc.
% (c) 2013 AR
%
% Steps to compile: latex latex latex
%
% For tracking purposes => this is v1.0 - Apr. 2013

\documentclass[letterpaper]{ar-1col}
\usepackage{showyourwork}
\usepackage[letterpaper]{geometry}

\usepackage{natbib}

\usepackage{graphbox}
\newcommand{\suz}[1]{\textcolor{magenta}{#1}}
\newcommand{\dan}[1]{\textcolor{green}{#1}}

\setcounter{secnumdepth}{4}
\usepackage{url}

% Metadata Information
\jname{Xxxx. Xxx. Xxx. Xxx.}
\jvol{AA}
\jyear{YYYY}
\doi{10.1146/((please add article doi))}

% macros
\newcommand{\boldsymbol}[1]{\mathbf{#1}}

\newcommand{\apjl}{Astrophysical Journal Letters}
\newcommand{\aj}{Astronomical Journal}
\newcommand{\apj}{Astrophysical Journal}
\newcommand{\apjs}{Astrophysical Journal Supplement}
\newcommand{\pasp}{Publications of the Astronomical Society of the Pacific}
\newcommand{\jgr}{Journal of Geophysical Research}
\newcommand{\aap}{Astronomy and Astrophysics}
\newcommand{\mnras}{Monthly Notices of the Royal Astronomical Society}
\newcommand{\actaa}{Acta Astronomica}
\newcommand{\nat}{Nature}
\newcommand{\prl}{Physical Review Letters}

% Document starts
\begin{document}

% Page header
\markboth{Aigrain \& Foreman-Mackey}{GPR for Astronomy}

% Title
\title{Gaussian Process regression for astronomical time-series}

%Authors, affiliations address.
\author{Suzanne Aigrain,$^1$ and Daniel Foreman-Mackey$^2$
  \affil{$^1$Department of Physics, University of Oxford, Oxford, UK, OX1 3RH; email: suzanne.aigrain@physics.ox.ac.uk}
  \affil{$^2$Center for Computational Astrophysics, Flatiron Institute, New York, USA, NY 10010}}

%Abstract
\begin{abstract}
  Abstract text, approximately 150 words.
\end{abstract}

%Keywords, etc.
\begin{keywords}
  keywords, separated by comma, no full stop, lowercase
\end{keywords}
\maketitle

%Table of Contents
\tableofcontents

\section{INTRODUCTION}
\label{sec:intro}

\begin{marginnote}[]
  \entry{GP}{Gaussian Process}
  \entry{GPR}{Gaussian Process Regression}
\end{marginnote}

Gaussian Processes (GPs) are a powerful class of statistical models, which allow us to define a probability distribution over random functions. Rather than write down an explicit mathematical formula for the function from which some observations are generated, we model the covariance between pairs of samples from the process, using our physical domain knowledge and/or available data to guide our modelling. GP Regression (GPR) has become increasingly popular in the astronomical community over the last decade. Part of the reason for this uptake is the growing availability and importance of time-domain datasets in astronomy. These systematically contain non-trivial random or unknown signals, whether astrophysical or instrumental, that need to be modelled. In many cases, these are nuisance signals, which we need to marginalise over in order to detect or measure other signals robustly. Sometimes, we are interested in the stochastic behaviour itself, and want to infer its characteristics or predict its behaviour. GPR offers a compelling solution: statistically principled, naturally Bayesian, and extremely flexible, yet mathematically simple. To new users, however, the lack of an explicit functional form for the model can make GPR can seem a little arcane. Furthermore, the computational cost of the method, which scales cubically with the dataset size, can be an obstacle. These factors initially impeded its dissemination in the astronomical community, but have been largely overcome in recent years thanks to the availability of user-friendly, computationally optimised software packages.

\subsection{Brief history}

First introduced in the geophysical literature as a spatial interpolation method by \citet{kriging}, GPR has since been adopted or re-invented in a wide range of other application domains. GPs were used in simulations in a wide range of astronomical sub-fields \citep[see e.g.][]{1980asfr.symp..159B,1988JGR....9311569C,1997ApJ...483L...1P}, but early mentions of GPs for modelling astronomical datasets \citep[see e.g.][]{1976MitAG..38..192D,1978A&A....70..777V,1991MGeo...16..313J} received limited attention.

To the best of our knowledge, the first use GPR in the refereed astronomical literature was published by \citet{prh92a} in the context of quasar variability, and for a long time this remained its main application domain in astronomy. GPR then gradually appeared in other areas, starting with photometric redshift estimation \citep{2006ApJ...647..102W}, then exoplanet transit observations \citep{2009ApJ...704...51C,2012MNRAS.419.2683G} and radial velocity planet searches \citep{2012MNRAS.419.3147A,2014MNRAS.443.2517H}. Nonetheless, GPR remained relatively niche and few astronomers had heard of it until a few years ago. To illustrate this point, we searched on the NASA Astrophysics Data System (ADS) for articles published in refereeed astronomy and astrophysics journals with the words `Gaussian Process' in the title or abstract. Until 2010, the number of such papers in any given year could be counted on the fingers of one hand. From that point onwards it has risen steadily, and in 2021 it was close to 100.

A number of important factors have contributed to recent the democratisation of GPR across a wide range of scientific disciplines, including the publication of a dedicated textbook \citep{gpml}, as well as the availability of cheap computing power and user-friendly, open-source GPR software. In astronomy, specifically, an additional factor has been at play: the rise of time domain surveys. Correlated noise in time-domain observations is a direct and unavoidable consequence of causation, and hence ubiquitous. Adequately modelling this correlated noise is vital when searching for faint signals, for example from exoplanets. Astrophysical sources, from accretion disks on all scales to magnetically active stars or cloudy brown darfs, also display complex, intrinsically or apparently stochastic behaviour, for which adequate modelling strategies are required. GPR is a natural choice to tackle these challenges. Rather than attempting to cover all applications of GPs to all of astrophysics, which would not be feasible in the space available, we have therefore opted to focus on its application to time-domain datasets in astronomy.

\subsection{Simulated examples}
\label{sec:sim_examples}

By way of motivation, before delving any further into the methodology, we have produced two simulated examples that illustrate the power of GPR for astronomical datasets. We have opted to use simulations rather than real data from the literature because knowning the ``ground truth'' allows us to quantify the performance of GPR against our metric of choice and to compare it to that of other methods.

\subsubsection{Lensed quasar light curves with time-delays}
\label{sec:sim_quasar}

In a nod to \citet{prh92a}, our first example consists of two simulated light curves of a hypothetical lensed quasar. Both light curves contain the same underlying signal, namely a single random draw from a GP prior with a Mat{\'e}rn $3/2$ kernel with parameters $A=0.1$ and $\lambda=5000\,{\rm d}^2$ (see Section \suz{XXX} for details \suz{check consistency of notation}). All GP calculations in this Section were performed using the {\sc george} package \citep{george}. Both light curves were simulated using the same set of times, which are drawn randomly in the interval $0$--$1500$\,d, but a time-delay of $\delta t=419.3$\,d was applied to the times for the second light curves before sampling the GP. In other words, the signal reaches the second source at a time $\delta t$ \textit{before} it reaches the first. We then added white Gaussian noise with standard deviation $\sigma_1=0.02$ and $\sigma_2=0.04$ vertical offsets of $M_1=1.0$ and $M_2=0.5$. The resulting simulated light curves are shown in the left panel of Figure~\ref{fig:sim_lcs}.

\begin{figure}[h]
  \centering
  \begin{minipage}[t]{0.48\linewidth}
    \includegraphics[width=\linewidth,vshift=1cm]{quasar_lc.pdf} 
  \end{minipage} \hfill
  \begin{minipage}[t]{0.48\linewidth}
    \includegraphics[width=\linewidth,vshift=1cm]{transit_lc.pdf}
  \end{minipage}
  \caption{Simulated datasets for the lensed quasar and transit examples. In the left panel, black dots and crosses with error bars show the light curves of the two quasar images, respectively, while the thin blue and red lines show the corresponding posterior samples for the simultaneous fit to both light curves using a Mat{\'e}rn $3/2$ GP.  In the right panel, black dots with error bars show the transit observations, while the thin blue and red lines lines show posterior samples for the models with white noise only and with a squared exponential GP, respectively. The posterior distributions for the models fit to both datasets are shown in Figure~\protect\ref{fig:sim_posteriors}. The {\sc Jupyter} notebooks used to produce this figure are available as Supplementary Material (\texttt{quasar\_time\_delay.ipynb} and \texttt{transit\_modelling.ipynb}).}
  \label{fig:sim_lcs}
\end{figure}

We then modelled both time-series simultaneously using, once again, a Mat{\'e}rn $3/2$ GP, varying the amplitude and length scale of the GP along with the means of the two time-series and the time-delay between them. We sampled the posterior using Markov Chain Monte Carlo (MCMC) with the {\sc emcee} package \citep{emcee} and broad uniform priors over all parameters. The resulting posteriors are shown Figure~\ref{fig:sim_posteriors}, while posterior samples are over-plotted on the data in Figure~\ref{fig:sim_lcs}. The true value of all parameters (including the light curve means, which aren't shown on the posterior plots as they are nuisance parameters) falls within the 68.5\% confidence interval. In particular, measured the time-delay is $\delta t = 418.9_{-1.3}^{+1.4}$\,d ($1$-$\sigma$).

\suz{If time permits: try downloading and modelling the actual data in \citet{1989A&A...215....1V} instead of simualted data?}

\dan{If time permits: rewrite using {\sc tinygp} instead of {\sc george} and {\sc emcee}?}

\subsubsection{Simulated transit example}
\label{sec:sim_transit}

\begin{figure}[h]
  \centering
  \begin{minipage}[t]{0.48\linewidth}
    \includegraphics[align=c,width=\linewidth,vshift=1cm]{quasar_posteriors.pdf} 
  \end{minipage} \hfill
  \begin{minipage}[t]{0.48\linewidth}
    \includegraphics[align=c,width=\linewidth,vshift=1cm]{transit_posteriors.pdf}
  \end{minipage}
  \caption{Corner plots for the fits to the two simulated quasar and transit light curves (left and right, respectively) shown in Figure~\ref{fig:sim_lcs}.  
  In each panel, the plots along the diagonal show 1-D histograms of the posterior samples for a subset of the fitted parameters, with vertical dashed lines indicating the 16$^{\rm th}$ and 84$^{\rm th}$ percentiles, while the off-diagonal plots show 2-D correlations between pairs of parameters. The horizontal and vertical solid black lines in each panel show the ``ground truth''. In the quasar case, we show the distribution for the GP variance $A$ and length scale $\lambda$ \suz{ensure notation consistent}, as well as for the time delay $\delta t$. In the transit case, the results of the white-noise only fit are shown in blue, while those of the GP fit are shown in red. These plots were generated using the {\sc corner.py} package \citep{corner}. \dan{How can we have them overlap with the transit one in the top right and the quasar one in the bottom left?} The {\sc Jupyter} notebooks used to produce this figure are available as Supplementary Material (\texttt{quasar\_time\_delay.ipynb} and \texttt{transit\_modelling.ipynb}).}
  \label{fig:sim_posteriors}
\end{figure}

For our second example, we simulate a transit observation containing correlated as well as white noise, and demonstrate the impact of modelling the correlated noise explicitly. The transit was simulated using the {\sc Pytransit} package \citep{2015MNRAS.450.3233P} with orbital period $P=1$\,d, planet-to-star radius ratio $R_{\rm p}/R_\star=0.1$, system scale $a/R_\star=3.0$, and impact parameter $b=0.1$. Conjunction occurs at $t=0$, and we simulated 100 equally spaced over the interval $-0.2\,{\rm d} \le t \le 0.2$\,d. We used the {\sc LDTk} package \citep{2015MNRAS.453.3821P} to include limb-darkening as expected for a Sun-like star observed in the bandpass of the Transiting Exoplanet Survey Satellite (TESS, \citealt{2015JATIS...1a4003R}). We then added both white Gaussian noise with a standard deviation of $\sigma=0.001$ and correlated noise simulated using a squared exponential GP with variance $A=0.000004$ and length scale $\lambda=0.0005\,{\rm d}^2$. The resulting light curve is shown in the right panel of Figure~\ref{fig:sim_lcs}.

We then used the same framework to model the simulated dataset, first ignoring the correlated noise and then accounting for it explicitly. As with the quasar example, all GP calculations were performed using {\sc george} and the posterior exploration was done using {\sc emcee}. Only $R_{\rm p}/R_\star$, $T_0$ and, where applicable, $A$ and $\lambda$ were allowed to vary (using broad, uninformative priors in all cases). The other parameters ($P$, $b$, $a/R_\star$ and the limb-darkening parameters) were fixed at their simulated values. In a real-world setting, these parameters would also be allowed to vary, but subject to priors derived from pre-existing observations of the transits and of the host star, which is beyond the scope of this simple example. The resulting posteriors for $R_{\rm p}/R_\star$ and $T_0$ are shown in Figure~\ref{fig:sim_posteriors}, while posterior samples from the two models are overplotted on the data in Figure~\ref{fig:sim_lcs}. Ignoring the correlated nature of the noise when modelling this dataset leads to significant biases in these parameters. On the other hand, when the correlated noise is accounted for using a GP, the posterior distributions are broader, but compatible with the ground truth. The posterior distributions for the GP parameters are not shown as they are nuisance parameters in this case, but they are also compatible with the true values.

\dan{If time permits switch to {\sc tinygp}}

\smallskip
The remainder of this review is structured as follows. Section~\ref{sec:basics} provides a brief, accessible introduction to the basic theory of GPR, and Section~\ref{sec:choices} discusses the key modelling choices one needs to make when using a GP to model data, with practical advice for how to make these choices. Section~\ref{sec:uses} gives an overview of applications of GPR to time-domain astronomical datasets to date. \suz{describe remaining sections}

% Heading 1
\section{BASICS OF GAUSSIAN PROCESS REGRESSION}
\label{sec:basics}

So far we have discussed how GPR has become widely used in astronomy and touched on some of their pros and cons, but we haven't explicitly defined what a GP is or explained how GPR works. This section gives a brief introduction to the theory of GPR, written in a way that we hope will be intelligible to most astronomers. For more details and practical exercises, we refer the reader to \citet{gpml} and \textcolor{red}{[Cite SA Saas Fee lecture notes if published]}.

\subsection{Formal definition}
\label{sec:def}

A GP is a type of \textit{stochastic process} based on the Gaussian probability distribution. A probability distribution describes a random variable with a finite number of dimensions. A stochastic process extends this concept to an infinite number of dimensions, allowing us to define a probability distribution over functions. Just what do we mean by ``extending to an infinite number of dimensions"? Well, this can be a little problematic mathematically, but we needn't worry about it, because in practice we only ever deal with finite samples from the stochastic process.

The formal definition of a GP is that the joint probability distribution over any finite sample $\mathbf{y} = (y_1, y_2, \ldots, y_n)^{\rm T}$ from the GP is a multi-variate Gaussian:
\begin{equation}
    p(\mathbf{y}) = \mathcal{N}(\mathbf{m}, \mathbf{K}),
\end{equation}
where $\mathbf{m}$ is the \emph{mean vector} and $\mathbf{K}$ the \emph{covariance matrix}.

The elements of the mean vector and covariance matrix are given by the \textit{mean function} $m$ and the \textit{covariance function} $k$, respectively:
\begin{equation}
    m_{i} = m(\boldsymbol{x}_i, \boldsymbol{\theta}),
\end{equation}
\begin{equation}
    K_{ij} = k(\boldsymbol{x}_i,\boldsymbol{x}_j, \boldsymbol{\phi}).
\end{equation}
where $\boldsymbol{x}_i$ is the set of inputs (independent variables) corresponding to the $i^{\rm th}$ sample. For time-series data, the inputs usually include, but aren't necessarily restricted to, the time $t_i$. The covariance function, also known as the \emph{kernel}, is the fundamental ingredient of a GP model, and considerable care must be taken to select it adequately (or to test different possibilities). Sometimes the mean function is assumed to be constant, or even zero, everywhere; this is often done in the wider GPR literature to keep derivations uncluttered. However, for many astrophysical applications where a GP is used to model a nuisance signal, the mean function contains the signal of interest and is important.

The parameters $\boldsymbol{\theta}$ and $\boldsymbol{\phi}$ of the mean and covariance function are known as the \textit{hyper-parameters} of the GP. Strictly speaking, the \textit{parameters} of the GP are the elements of the mean vector and covariance matrix. However, for a given set of inputs and choice of mean and covariance functions, these numerous individual parameters are entirely determined by the the hyper-parameters, so the user interacts with the latter in much the same way as they would with the parameters of a more classical model.

Although is is possible to construct and use stochastic process models based on other distributions, GPs are by far the most popular, for two main reasons. The first is Central Limit theorem: it implies that the assumption of Gaussianity is often at least approximately correct. The second is that Gaussian distributions obey simple mathematical identities for marginalisation and conditioning, that enable inference with GPs to be performed analytically, as we will see later.

\subsection{From least-squares to GPR}
\label{sec:from_lsq}

The formal definition above does not necessarily provide an intuitive understanding of how GPs work. Most readers of this review will, however, be comfortable with least-squares regression. In this section we will show that GPR can be thought of as a generalisation of least-squares regression, allowing for correlated noise (or signals) in the data. Conversely, least-squares regression is a special case of GPR, where the covariance matrix is assumed to be purely diagonal, and the variances associated with each observation are known \textit{a priori}.

\subsubsection{Revsisiting least squares}

Consider $N$ observations of a variable $\mathbf{y} = \{ y_i\}_{i=1, \ldots, N}$, taken at times $\mathbf{t} = \{ t_i\}$, with associated measurement uncertainties $\boldsymbol{\sigma}=\{\sigma_i\}$. We wish to compare these to a model function $m(t,\boldsymbol{\theta})$ controlled by parameters $\boldsymbol{\theta}=\{\theta_j\}_{j=1,\ldots,M}$. In most astrophysical applications, we are interested in estimating some or all of those parameters. In least-squares regression, we minimize the quantity 
\begin{equation}
  \chi^2 \equiv\sum_{i=1}^N (y_i-m_i)^2/\sigma^2_i,
\end{equation}
where $m_i \equiv m(x_i,\boldsymbol{\theta})$, with respect to $\boldsymbol{\theta}$. Where does this come from?

Let us assume that the observations are given by 
\begin{equation}
  y_i=m(t_i,\boldsymbol{\theta})+\epsilon_i, 
\end{equation}
where $\epsilon_i$ is the measurement error, or noise, on the $i^{\rm th}$ observation. Furthermore, let us assume that $\epsilon_i$ is drawn from a Gaussian distribution with mean $0$ and variance $\sigma_i^2$:
\begin{equation}
  p(\epsilon_i)=\mathcal{N}(0,\sigma_i^2) \equiv \frac{1}{\sqrt{2\pi} \sigma_i} \exp\left(-\frac{\epsilon_i^2}{2\sigma_i^2} \right),
\end{equation}
then the \emph{likelihood} for the $i^{\rm th}$ observation is simply
\begin{equation}
  \mathcal{L}_i\equiv p(y_i|\boldsymbol{\theta})=\mathcal{N}(m_i,\sigma_i^2) =\frac{1}{\sqrt{2\pi} \sigma_i} \exp\left[-\frac{(y_i-m_i)^2}{2\sigma_i^2} \right].
\end{equation}
Furthermore, we also assume that the noise is uncorrelated, or white, meaning that the $\epsilon_i$'s are drawn independently from each other from their respective distributions. Then, the likelihood for the whole dataset $\mathbf{y}$ is merely the product of the likelihoods for the individual observations:
\begin{equation}
  \mathcal{L} \equiv p(\mathbf{y}|\boldsymbol{\theta}) = \prod_{i=1}^N \mathcal{L}_i = \prod_{i=1}^N \left\{ \frac{1}{\sqrt{2\pi} \sigma_i}
  \exp\left[-\frac{(y_i-m_i)^2}{2\sigma_i^2} \right] \right\},
\end{equation}
where $\mathbf{m}=\{m_i\}_{i=1,\ldots,N}$. From this one can readily see that $\ln \mathcal{L} = \mathrm{constant} - 0.5 \chi^2$, where the constant depends only on the $\sigma$'s. Thus, if the $\sigma$'s are known, maximizing $\mathcal{L}$ is equivalent to minimizing $\chi^2$. In other words, least-squares regression yields the Maximum Likelihood Estimate (MLE) of the parameters under the assumption of white, Gaussian noise with known variance.

\subsubsection{Link to Gaussian Process Regression}
Let us now re-write the likelihood in matrix form:
\begin{equation}
\label{eq:gp_like}
  \mathcal{L} = \left(\frac{1}{2\pi} \right)^{N/2}
  \frac{1}{|\mathbf{K}|} \exp \left(-(\mathbf{y}-\mathbf{m})^{\mathrm{T}} \mathbf{K}^{-1}
  (\mathbf{y}-\mathbf{m}) \right),
\end{equation}
where $\mathbf{K}$ is a purely diagonal ($N$, $N$) matrix with elements $K_{ij} = \delta_{ij} \sigma_i^2$ ($\delta_{ij}$ being the discrete Kronecker delta function). This is, of course, the \emph{covariance matrix} of the model.

Now, instead of assuming that the covariance matrix takes this very specific form, let us allow a more flexible model for the covariance:
\begin{equation}
  K_{ij} = k(t_i,t_j,\boldsymbol{\phi}) + \delta_{ij} \sigma_i^2,
\end{equation}
where $k$ is a \emph{covariance function} or \emph{kernel}, controlled by parameters $\boldsymbol{\phi}$. The result is a GP, and its likelihood is still given by Equation~(\ref{eq:gp_like}). Depending on the choice of kernel and parameters, the covariance matrix can now have non-zero off-diagonal elements, allowing us to explicitly model correlated noise or stochastic signals in the data. 
Note that the likelihood depends not only on the argument of the exponential, but also on the determinant of the covariance matrix, which therefore needs to be evaluated explicitly. This term acts as a built-in Occam's razor, automatically penalising more complex models.

The kernel $k$ encodes our beliefs about the stochastic, or random, element of the model, in just the same way as the mean function $m$ encodes our beliefs about the deterministic component of the model. For example, in many circumstances, we would expect that two observations taken close together in time should be more strongly correlated than observations taken further apart, and we would use a decreasing function of the time interval $\tau=|t_i-t_j|$ to represent that. One of the most widely used covariance functions is the squared exponential 
\begin{equation}
    k_{\rm SE} = A \exp{-\tau^2/\lambda}.
\end{equation}
This gives rise to smooth random functions with variance $A$ and characteristic length-scale $\sqrt{\lambda}$. We present other commonly used kernels and discuss how to select one in Section~\ref{sec:kernels}.

\subsection{Inference with a GP}

\begin{figure}[h]
  \includegraphics[width=\linewidth]{GPR_workflow.pdf}
  \caption{Schematic of the workflow for Bayesian inference with a GP. The process starts with getting some data (blue box), and making some choices about how we are going to model it (grey boxes, these choices are discussed in Section~\ref{sec:choices}). The mathematical framework presented in this section can then be used (green boxes) to evaluate the likelihood and posterior distribution over the HPs which can be optimized or sampled. We can also condition the GP on the data to predict observations at locations where we do not yet have observations. \suz{Figure will be updated with nicer version.}}
  \label{fig:GPRworkflow}
\end{figure}

Now that we know how to evaluate the likelihood (Equation~\ref{eq:gp_like}), we are ready to perform \textit{inference}, that is to use observations to update our prior beliefs about the system we are observing. The overall Bayesian inference workflow for GPR regression is illustrated schematically in Figure~\ref{fig:GPRworkflow}. For example, we can optimize the likelihood with respect to the hyper-parameters $\boldsymbol{\phi}$ and $\boldsymbol{\theta}$. This is called \emph{training} the GP. In doing so, we are learning the properties of the correlated signal from the data, hence GPR is a machine learning method.

Given adequate priors, we can also evaluate the \emph{posterior} distribution:
\begin{equation}
  p(\boldsymbol{\theta},\boldsymbol{\phi}|\mathbf{y})=\frac{p(\mathbf{y}|\boldsymbol{\theta},\boldsymbol{\phi})p(\boldsymbol{\theta},\boldsymbol{\phi})}{p(\mathbf{y})},
\end{equation}
where $p(\mathbf{y})=\int p(\mathbf{y}|\boldsymbol{\theta},\boldsymbol{\phi})p(\boldsymbol{\theta},\boldsymbol{\phi}) \mathrm{d}\boldsymbol{\theta} \mathrm{d}\boldsymbol{\phi}$.
Usually, the posterior cannot be evaluated analytically, so we resort to sampling methods such as Markov Chain Monte Carlo (MCMC) or nested sampling.

In many astrophysical applications of GPs, we wish to take into account correlated noise, but we are not interested in the noise \emph{per se}, only on the impact it has on the parameters of the mean function, $\boldsymbol{\theta}$. We thus \emph{marginalize} (integrate over) the \emph{nuisance parameters} $\boldsymbol{\phi}$.

\subsection{Making predictions}
\label{sec:pred}

An important use case for GPR is as a statistically principled interpolation method. We wish to ``learn'' an unknown function that gave rise to some data, in order to make predictions for some new set of of inputs. Importantly, a GP model provides not a point estimate but a full probability distribution for the function at any desired location(s) in the input domain. This allows for robust uncertainty propagation (though there are some important caveats we will touch upon), and can also motivate strategies for active sampling (deciding when or where to make new observations).

\subsubsection{The predictive equations}

Given some existing observations $\mathbf{y}$, taken at times $\mathbf{t}$, how do we make predictions at some new set of times $\mathbf{t}_\star$? We are after $p(\mathbf{y}_\star|\mathbf{y})$, the \textbf{conditional} probability distribution for $\mathbf{y}_\star$ given $\mathbf{y}$. In GPR this is also known as the \emph{predictive distribution}, because it is often used to extrapolate a time-series dataset forwards. The `magic' of GPR is that the predictive distribution is also Gaussian, and its mean and covariance are given by simple analytic relations:
\begin{equation}
  \label{eq:pred}
  \mathbf{f}_\star = \mathbf{m}_\star + \mathbf{K}_\star^{\mathrm{T}} \, \mathbf{K}^{-1} \, (\mathbf{y} - \mathbf{m}) ~\mathrm{~~~and~~~}
  \mathbf{C}_\star =   \mathbf{K}_{\star\star} - \mathbf{K}_\star^{\mathrm{T}} \, \mathbf{K}^{-1} \, \mathbf{K}_\star,
\end{equation}
where we have introduced the notation $\mathbf{m}_\star \equiv m(\mathbf{t}_\star,\boldsymbol{\theta})$, $\mathbf{K}_\star \equiv k(\mathbf{t},\mathbf{t}_\star, \boldsymbol{\phi})$ and
$\mathbf{K}_{\star\star} \equiv k(\mathbf{t}_\star,\mathbf{t}_\star,\boldsymbol{\phi})$. Note that, as real observations are always noisy, $\mathbf{K}$ generally includes a white noise term ($\delta_{ij} \sigma^2_i$), but $\mathbf{K}_\star$ does not. Depending on whether one wishes the predictive variance (the diagonal elements of $\mathbf{C}_\star$)  to account for measurement uncertainties or not, one can include the white noise term in $\mathbf{K}_{\star\star}$ or leave it out.

\subsubsection{Properties of the  predictive distribution}
A closer look at Equations~(\ref{eq:pred}) reveals some important properties.
First, a GP is a \textit{linear predictor}. The predictive mean for a specific time $t_\star$ can be written as a linear combination of the observations:
$f_\star = \mathbf{w}^{\mathrm{T}} \, \mathbf{y}$, where $\mathbf{w} = \mathbf{K}^{-1} \, \mathbf{k}_\star$ and $\mathbf{k}_\star=k(\mathbf{t},t_\star,\phi)$. It can also be written as
a linear combination of covariance functions centred on the training points:
$f_\star = \mathbf{\alpha}^\mathrm{T} \,  \mathbf{k}_\star$ where $\mathbf{\alpha} = \mathbf{K}^{-1} \, \mathbf{y}$.
These linearity properties are very important to understand the behaviour of GPs. The shed light on the relationship between GPR and standard linear models with very large numbers of free parameters \citep[see e.g.][for a more detailed discussion]{2021PASP..133i3001H}, 
and lead to a number of powerful extensions that are beyond  scope of this review.

Second, the predictive covariance is independent of the data. $\mathbf{C}_\star$ depends only on the \emph{locations} $\mathbf{t}$ of the observations, not on their values $\mathbf{y}$. This has important consequences for observation planning: if we know the covariance function $k$ and its parameters, we can decide when to observations to optimise our predictions at a given (set of) time(s). However, in practice we rarely know the hyper-parameters \textit{a priori}, and the predictive posterior distribution marginalised over the hyper-parameters \textit{does} depend on the observations.

Finally, as $\mathbf{K}$ is positive semi-definite, so is $\mathbf{K}^{-1}$. Therefore, $\mathbf{k}_\star^{\mathrm{T}} \, \mathbf{K}^{-1} \, \mathbf{k}_\star \ge 0,$ and $\mathrm{Var}(\mathbf{f}_\star) \leq k(t_\star,t_\star)$. This is as we would expect: obtaining more data should only ever improve the accuracy of our predictions.

\subsubsection{Cautionary notes}

It is important to note that the behaviour of $\mathbf{f}_\star$ is \emph{not} the same as that of individual samples from the predictive distribution. Typically,  $\mathbf{f}_\star$ tends to be smoother than individual samples. This should be borne in mind when displaying GPR results or using them in subsequent analysis.

It is also important to note that the predictive variance accounts for the imperfect ability of the specific model under consideration to explain the data, but not for the choice of model (i.e. the choice of mean and kernel functions and their parameters). We discuss how to choose a kernel function and fit for its parameters in the next section.

\section{GAUSSIAN PROCESS MODELING DECISIONS}
\label{sec:choices}

% Assigned: DFM

% \paragraph{Kernels} Selecting a kernel function is how the user builds physical domain knowledge into the Gaussian process model. In this section, we will both try to dispel some common misconceptions about how to choose covariance functions and their parameters, and provide some practical guidelines for how to make these modelling choices in practice, highlighting specific examples from the astronomical literature. Estimated length: 5-10 pages

% Question: will this also be the section where the most commonly used kernel functions in astrophysics are introduced or will that be part of section1? This will affect the relative length of the two sections, but overall I'd expect it to be 15-20 pages for both sections

% \paragraph{Tuning} This section will cover the process of "training" a Gaussian process model, i.e. tuning the parameters of the covariance function and, if included, the mean function. Compared to standard texts from other fields, this topic will be covered in more depth because parameter inference (as opposed to interpolation or prediction) is one of the most common applications of Gaussian processes in astronomy. Once again we will aim to dispel some misconceptions and provide practical advice. This will also be where we illustrate how a Gaussian process model can be embedded in a Bayesian inference framework using posterior sampling, and point out the importance of marginalising over nuisance parameters. Estimated length: 5-10 pages

In the previous section, we introduced the GP sampling distribution, and results of conditioning a GP on observed data.
A crucial part of such a GP model is it's ``kernel'' or ``covariance'' function---called $k(t_i,\,t_j;\,\phi)$ above---which specifies the covariance between two input coordinates $t_i$ and $t_j$.
Throughout the astronomical literature, GP modeling generally includes a non-trivial mean function.
However, in other fields, this mean is regularly omitted, meaning that the kernel function fully specifies the behavior of the GP model, making this choice \emph{the fundamental GP modeling decision}.
That being said, in our experience, new users of GP models will often focus and worry more than necessary about the choice of kernel function for their problem.
As with any probabilistic modeling problem, there are several well-defined workflows for motivating, selecting, and validating your choice of kernel.
In this section we walk through this process in detail.

% In this section, we formally define this kernel function and the properties it must have, we provide some examples and advice for setting up GP models, and discuss the most commonly used inference schemes for these models.

% DFM to do
% More conceptual intro; requirements
% Stationary vs non-stationary, Fourier repr; some examples
% kernel design: sums, products, etc.
% hyperparameter tuning / marginalization

\subsection{Gaussian Process Covariance Functions}
\label{sec:kernels}

\begin{marginnote}[]
  \entry{Kernel function}{The ``kernel'' or ``covariance'' function, typeset here as $k(t_i,\,t_j;\,\phi)$ is a parametric description of the covariance between two data points $t_i$ and $t_j$. The parameter $\phi$ indicates that these functions are typically parameterized by a set of parameters, typically referred to as ``hyperparameters''.}
\end{marginnote}

\begin{marginnote}[]
  \entry{Hyperparameters}{In a GP model, the kernel function is typically parameterized by a set of parameters that we will label as $\phi$ and refer to as ``hyperparameters''. This label is used to distinguish these parameters from the parameters of the mean function, which we will refer to as $\theta$.}
\end{marginnote}

The kernel function can be any positive semi-definite scalar function, but some will be more useful than others.
A commonly (over-)used kernel function is the ``exponential-squared'' or ``radial basis function'' kernel
\begin{equation}
  k(t_i,\,t_j;\,\phi) = \sigma^2\,\exp\left[-\frac{(t_i - t_j)^2}{2\,\tau^2}\right]
\end{equation}
where the ``hyperparameters'' $\phi$ are $\sigma$ and $\tau$.
In this case, the hyperparameters specify the typical amplitude $\sigma$ and timescale $\tau$ of the variance of the process.
SOMEFIGURE shows how changing these parameters affect our GP model.
